Approximation as Differentiation

### 2025-06-18 main/first-order.tex
Tweak.

The families construction, regarded as a covariant 2-functor $\Fam(-): \Cat \to \Cat$, sends functors $F:
\cat{C} \to \cat{D}$ to functors $\Fam(F): \Fam(\cat{C}) \to \Fam(\cat{D})$ that reassign the target category
by postcomposing with $F$, sending objects $(X, \partial X)$ to $(X, F \comp \partial X)$ and morphisms $(f,
\partial f)$ to $(f, F \comp \partial f)$, where $F \comp \partial f$ denotes the natural transformation where
$(F \comp \partial f)(x) = F(\partial f(x))$. If $F$ happens to be a (strong) monad on $\cat{C}$, then
$\Fam(F)$ is a (strong) monad on $\Fam(\cat{C})$, computed by lifting the original monad $F: \cat{C} \to
\cat{C}$ pointwise over families.

@@

One might also like to allow the programmer to embed additional ``approximation points'' into composite
structures, by providing them with a \emph{lifting} monad $\synVar{L}$. Rather than the usual definition of
lists of natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could then (within the language)
define a type of ``approximation lists'' $\mu\alpha.\synVar{L}\,(1 + \synVar{L}\,\synVar{nat} \times \alpha)$,
along with an approximating version of multiplication $*_{\text{appr}}: \synVar{L}\,\synVar{nat} \times
\synVar{L}\,\synVar{nat} \to \synVar{L}\,\synVar{nat}$, expressible as
\[x *_{\text{appr}} y = \tmBind{x}{\tmFun{x'}{\syn{if}\;x' \equiv
0\;\syn{then}\;\tmReturn{0}\;\syn{else}\;\tmBind{y}{\tmFun{y'}{\tmReturn{x' * y'}}}}}\].

\noindent This would allow the programmer to emulate the short-circuiting behaviour of $\mathrm{strictOr}$
from \exref{strict-short-circuit}, or the short-circuiting multiplication used in \citet{perera22} for slicing
matrix convolutions, without baking this specific choice of approximation into the interpretation of
individual operators.

\todo{Show how $\LatGal$ has a lifting monad?}

@@

We can verify that $\tangents$ is functorial, regardless of whether the linear maps $\RR^m \linearto \RR^n$
happen to be derivatives. But if $f_2(x)$ is the derivative of $f_1$ at $x$ for any $x \in \RR^m$ and
similarly for $g$ then the derivatives compose according to the forward chain rule and $(g \comp f)_2(x)$ is
the derivative of $(g \comp f)_1$ at $x$. Thus this setup provides an algorithmic method for computing the
forward derivative of a composite function, without have to recompute the function itself, as long as all
atomic operations provide derivatives.

>>>>

@@

One might also like to allow the programmer to embed additional ``approximation points'' into composite
structures, by providing them with a \emph{lifting} monad $\synVar{L}$. Rather than the usual definition of
lists of natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could then (within the language)
define a type of ``approximation lists'' $\mu\alpha.\synVar{L}\,(1 + \synVar{L}\,\synVar{nat} \times \alpha)$,
or perhaps a custom approximating version of multiplication $*_{\text{appr}}: \synVar{L}\,\synVar{nat} \times
\synVar{L}\,\synVar{nat} \to \synVar{L}\,\synVar{nat}$, expressible as
\[x *_{\text{appr}} y = \tmBind{x}{\tmFun{x'}{\syn{if}\;x' \equiv
0\;\syn{then}\;\tmReturn{0}\;\syn{else}\;\tmBind{y}{\tmFun{y'}{\tmReturn{x' * y'}}}}}\]

\noindent This would allow the programmer to emulate the short-circuiting behaviour of $\mathrm{strictOr}$
from \exref{strict-short-circuit}, or the short-circuiting multiplication used in \citet{perera22} for slicing
matrix convolutions, without baking specific choices of approximation into the interpretation of individual
operators.

To interpret the syntactic type constructor $\synVar{L}$, \todo{show how $\LatGal$ has a lifting monad.}

The families construction, regarded as a covariant 2-functor $\Fam(-): \Cat \to \Cat$, sends functors $F:
\cat{C} \to \cat{D}$ to functors $\Fam(F): \Fam(\cat{C}) \to \Fam(\cat{D})$ that reassign the target category
by postcomposing with $F$, sending objects $(X, \partial X)$ to $(X, F \comp \partial X)$ and morphisms $(f,
\partial f)$ to $(f, F \comp \partial f)$, where $F \comp \partial f$ denotes the natural transformation where
$(F \comp \partial f)(x) = F(\partial f(x))$. If $F$ happens to be a (strong) monad on $\cat{C}$, then
$\Fam(F)$ is a (strong) monad on $\Fam(\cat{C})$, computed by extending the original monad $F: \cat{C} \to
\cat{C}$ componentwise to families. Thus the lifting monad $\mathcal{L}$ in $\LatGal$ is also a lifting
monad in $\Fam(\LatGal)$.

@@

We can verify that $\tangents$ is functorial, regardless of whether the linear maps $\RR^m \linearto \RR^n$
happen to be derivatives. But if $f_2(x)$ is the derivative of $f_1$ at $x$ for any $x \in \RR^m$ and
similarly for $g$ then the derivatives compose according to the forward chain rule and $(g \comp f)_2(x)$ is
the derivative of $(g \comp f)_1$ at $x$. Thus this setup provides an algorithmic method for computing the
forward derivative of a composite function, as long as all atomic operations provide derivatives, without have
to recompute the function itself.
